Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-15 response_letter.txt

########################################
MAJOR open issues:

* Emphasize: we can do pseudoknots (like Blueprint)!
  already in abstract by now

* improve optimization by connected components of the compatibility graph

* simplify
   -- (sequence) positions/nucleotides instead of variables/values
   -- (partial) sequences instead of (partial) assignments
   -- evaluation of functions under assignments; maybe provide example---rethink if
      the abstraction is necessary (especially if separator to parent is always size 1)
   -- potentially simplify since separator to parent is always size 1

* TODO discuss (for the question of artificial vs. natural / reviewer 3):
  -- Adenin-Riboswitch (Schwalbe). Details aus Stefans Diss: 3-state/4-state riboswitch.
     Experimentally verified designs were done only for 2-state.

* RESULTS
    -- three additional pseudoknotsets (for Table 1). 
    -- linear correlation plot for Turner-energies fitted to simple energies (stacking?) 
         (additionally with PKs!?)
    -- shifting of targets (Figure 2) with PKs
    -- time benchmark; run time for 1000 instances

########################################

@@

-Page 6 "requires O(..." + "time"
DONE

-Page 7 For Theorem 1, I was wondering whether it is really necessary to
consider the total number of connected components. It seems that only the
maximum number of components per bag is crucial?

TODO: obviously, we confused the reviewer---
MOREOVER: strategy should be optimized. Instead of fixing all polarities in the beginning,
it suffices to fix the polarities simultaneously in one bag....

TODO: rewrite and present a better optimization strategy due to the base pairing constraints (compatibility)

-Page 7 "we encode the network energy ..." I was a bit confused here. What
is w_l? Is it possible to give the formula?
DONE

@@

Controling the energy barrier between two stable structure An important
feature when designing multi-stable RNA

>>>>> INTERESTING IDEA; rather for design than sampling. We could try to optimize 
the direct path barrier in this way (for producing decoys...)

* It was not clear to me if the authors implemented dangling ends?

>>>>> NO, but it is covered by the framework; if possible, reformulate the
resepctive sentence CHECK

### 2018-02-19 RNARedPrint-Bioinfo2018.tex
Replace term 'compatibility graph' by 'base pair dependency graph'

\begin{figure*}[t]
{\centering\includegraphics[width=.8\textwidth]{Figs/Workflow}\\}
\caption{General outline of \ourprog{} for base pair-based energy models. From a set of target secondary structures (i), base-pairs are merged (ii) into a compatibility graph (iii) and transformed into a tree decomposition (iv). The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v). An adaptive scheme learns weights to achieve targeted energies and \GCb-content, leading to the production of suitable designs (vi).}
\label{fig:workflow}
\end{figure*}

@@

We consider a fixed set of target RNA structures
$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
\Def{compatibility graph} $G_{\R}$ with nodes $\{1,\dots,n\}$ and
edges $S\bigcup_{\ell\in[1,k]} R_\ell$. While we typically consider
additional dependencies (in the dependency graph), the edges in the
compatibility graph pose constraints on the possible bases due to
canonical base pairing, which are present in all considered settings.

@@

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$. Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$. As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides. The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.

@@

     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\;
     $p \gets p\ \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
%
%     $p \gets 1$\;
%     \For{$f\in \phi(v)$}{
%  	      $p \gets p \cdot \exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$\; 
%     }
%     \For{$v\to{}w\in T$}{
%  	      $p \gets p \cdot \evalfor{\Message{v}{w}}{\substitute{\val}{\val'}}$\; 
%     }
  	  $r \gets r - p$\; 
  	  \If{$r<0$}{
  	    $a_{\rm res} \gets \substitute{\val}{\val'}$\;
  	  \For{$(v\to{}w)\in T$}{
  	      $a_{\rm res} \gets a_{\rm res} \mid \Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)$\; 
     }
  	  \Return {$a_{\rm res}$}
  	  \;}
 }
 }
 \caption{Stochastic backtrack algorithm for assignments in the Boltzmann distribution.}\label{alg:sampling}
\end{algorithm}
%For example, for design to a single non-crossing target in the nearest neighbor energy model, the complexity of Algs.~\ref{alg:pf} and \ref{alg:sampling} degrades to linear time (as reported for IncaRNAtion), since there are only $O(n)$ many functions and nodes (in the sequence length $n$). Moreover, the dependencies are tree-like owing to the tree-like non-crossing structure; this implies a constantly bounded maximum treewidth.

As already noted by \citet{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component of the compatibility graph, such that the variables of one
partition are restricted to values $\{\Ab,\Gb\}$, but for other
partition, to values $\{\Cb,\Ub\}$. This suggests a strong
optimization of our algorithms, which reduces the basis in the
exponential factor from 4 to 2. The price for this reduction is a
binary choice for each of $c$ connected component. For each of the
$2^c$ combinations, the partition functions can be computed separately
in $O(n\,m\, 2^{w+1})$ time. This strategy is applied only if
$c< w+1$. 
%% SW: This could be optimized further by applying the strategy only to (optimally) selected components
Finally, we compute the sums of the partition functions for all possible combinations.

\begin{theorem}[Complexities]
  For sequence length $n$, $k$ target structures, treewidth $w$ and
  a compatibility graph having $c$ connected components, $t$ sequences
  are sampled from the Boltzmann distribution  in
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  $\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}

@@

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special compatibility graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see \citet{Bender1983} and
\citet{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb-content of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\citep{Reinharz2013}.

@@

Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps. At each step, we resampled (uniformly at random) the positions of a randomly selected component in the compatibility graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
%
Once again, for all instances, we observe a positive improvement of the quality of Boltzmann designs over uniform ones, suggesting a superior potential for optimization ($\sim 32\%$ avg improvement, Table~\ref{tab:benchmark-results}, optimized). This improvement is however more modest than in the absence of optimization, but we attribute this relative decay to the uniform nature of the resampling strategy, which somehow ends up canceling the benefits of the initial Boltzmann distribution, by uniformizing the sequence over longer runs. This suggests an alternative resampling strategy based on Boltzmann-distributed steps.

@@

In addition, we generated four sets of instances of increasing
complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
respectively specify 3,4,5, and 6 target structures for sequence
length 100.  For each instance (100 instances per set), we generated a
set of $k$ ($k=3,\dots,6$) compatible structures as follows
\begin{itemize}
\item Generate a random sequence of length 100;
\item Compute its minimum free energy structure (ViennaRNA package);
\item Add the new structure to the instances if the resulting compatibility graph is bipartite;
\item Repeat until $k$ structures are collected.
\end{itemize}
For each instance, we generated the dependency graphs in the base pair
model and in the stacking model. Then, we performed tree decomposition
(using strategy ``GreedyFillIn'' of LibTW~\citep{Dijk2006}) on each dependency
graph. The obtained treewidths are reported in
Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
corresponding run-times of the tree decomposition.
Fig.~\ref{fig:td-example} shows the tree decompositions for an example
instance from set RF3.

>>>>

\begin{figure*}[t]
{\centering\includegraphics[width=.8\textwidth]{Figs/Workflow}\\}
\caption{General outline of \ourprog{} for base pair-based energy models. From a set of target secondary structures (i), base-pairs are merged (ii) into a (base pair) dependency graph (iii) and transformed into a tree decomposition (iv). The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v). An adaptive scheme learns weights to achieve targeted energies and \GCb-content, leading to the production of suitable designs (vi).}
\label{fig:workflow}
\end{figure*}

@@

We consider a fixed set of target RNA structures
$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
\Def{base pair dependency graph} $G_{\R}$ with nodes $\{1,\dots,n\}$
and edges $S\bigcup_{\ell\in[1,k]} R_\ell$, which describe the minimal
dependencies present in all relevant settings due to the requirement
of canonical base pairing.

@@

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given base pair dependency graph $G=(V,E)$. Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$. As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its base pair dependency graph is bipartite, which can be tested in linear time.
Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides. The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.

@@

     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\;
     $p \gets p\ \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
%
%     $p \gets 1$\;
%     \For{$f\in \phi(v)$}{
%  	      $p \gets p \cdot \exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$\; 
%     }
%     \For{$v\to{}w\in T$}{
%  	      $p \gets p \cdot \evalfor{\Message{v}{w}}{\substitute{\val}{\val'}}$\; 
%     }
  	  $r \gets r - p$\; 
  	  \If{$r<0$}{
  	    $a_{\rm res} \gets \substitute{\val}{\val'}$\;
  	  \For{$(v\to{}w)\in T$}{
  	      $a_{\rm res} \gets \substitute{a_{\rm res}}{\Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)}$\; 
     }
  	  \Return {$a_{\rm res}$}
  	  \;}
 }
 }
 \caption{Stochastic backtrack algorithm for assignments in the Boltzmann distribution.}\label{alg:sampling}
\end{algorithm}
%For example, for design to a single non-crossing target in the nearest neighbor energy model, the complexity of Algs.~\ref{alg:pf} and \ref{alg:sampling} degrades to linear time (as reported for IncaRNAtion), since there are only $O(n)$ many functions and nodes (in the sequence length $n$). Moreover, the dependencies are tree-like owing to the tree-like non-crossing structure; this implies a constantly bounded maximum treewidth.

As already noted by \citet{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component of the base pair dependency graph, such that the variables of one
partition are restricted to values $\{\Ab,\Gb\}$, but for other
partition, to values $\{\Cb,\Ub\}$. This suggests a strong
optimization of our algorithms, which reduces the basis in the
exponential factor from 4 to 2. The price for this reduction is a
binary choice for each of $c$ connected component. For each of the
$2^c$ combinations, the partition functions can be computed separately
in $O(n\,m\, 2^{w+1})$ time. This strategy is applied only if
$c< w+1$. 
%% SW: This could be optimized further by applying the strategy only to (optimally) selected components
Finally, we compute the sums of the partition functions for all possible combinations.

\begin{theorem}[Complexities]
  For sequence length $n$, $k$ target structures, treewidth $w$ and
  a base pair dependency graph having $c$ connected components, $t$ sequences
  are sampled from the Boltzmann distribution  in
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  $\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}

@@

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special base pair dependency graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see \citet{Bender1983} and
\citet{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb-content of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\citep{Reinharz2013}.

@@

Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps. At each step, we resampled (uniformly at random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
%
Once again, for all instances, we observe a positive improvement of the quality of Boltzmann designs over uniform ones, suggesting a superior potential for optimization ($\sim 32\%$ avg improvement, Table~\ref{tab:benchmark-results}, optimized). This improvement is however more modest than in the absence of optimization, but we attribute this relative decay to the uniform nature of the resampling strategy, which somehow ends up canceling the benefits of the initial Boltzmann distribution, by uniformizing the sequence over longer runs. This suggests an alternative resampling strategy based on Boltzmann-distributed steps.

@@

In addition, we generated four sets of instances of increasing
complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
respectively specify 3,4,5, and 6 target structures for sequence
length 100.  For each instance (100 instances per set), we generated a
set of $k$ ($k=3,\dots,6$) compatible structures as follows
\begin{itemize}
\item Generate a random sequence of length 100;
\item Compute its minimum free energy structure (ViennaRNA package);
\item Add the new structure to the instances if the resulting base pair dependency graph is bipartite;
\item Repeat until $k$ structures are collected.
\end{itemize}
For each instance, we generated the dependency graphs in the base pair
model and in the stacking model. Then, we performed tree decomposition
(using strategy ``GreedyFillIn'' of LibTW~\citep{Dijk2006}) on each dependency
graph. The obtained treewidths are reported in
Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
corresponding run-times of the tree decomposition.
Fig.~\ref{fig:td-example} shows the tree decompositions for an example
instance from set RF3.
